k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2024-03-06 paper.Rmd
abstract

    Numerous applications in psychological research require that a data set is partitioned via the maximization---instead of the more common minimization---of a clustering criterion to obtain high similarity between groups and heterogeneity within groups. Such anticlustering may consider criteria of overall between-group similarity and/or pairwise within-group dissimilarity. Brusco et al. (2020) proposed a bicriterion algorithm (BILS) to obtain partitionings that simultaneously maximize the diversity, corresponding to between-group similarity, and the dispersion, corresponding to pairwise within-group dissimilarity. As a heuristic, however, the BILS does not necessarily identify globally optimal criterion values. We provide four extensions to the bicriterion approach for anticlustering: First, we present a novel exact algorithm that returns multiple globally optimal partitions with regard to the dispersion. Despite its theoretical computational hardness, our method scales to rather large data sets (e.g., $N > 5000$ for $K = 3$ groups). Second, by extending existing optimal and heuristic approaches (including BILS), we evaluate how to best obtain a high diversity on top of an optimal dispersion provided by our algorithm. Third, we allow that the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices. Forth, by reformulating the k-means clustering criterion as an average diversity criterion, we allow that other anticlustering objectives of between-group similarity can be optimized alongside the dispersion instead of only the diversity. The original BILS as well as our extensions are available via the free and open source R package `anticlust`. Practical examples are included to illustrate their application.

>>>>

    Numerous applications in psychological research require that a data set is partitioned via the maximization---instead of the more common minimization---of a clustering criterion. Such anticlustering seeks for high similarity between groups (i.e., maximum diversity) and/or high pairwise dissimilarity within groups (i.e., maximum dispersion). Brusco et al. (2020) proposed a bicriterion algorithm (BILS) that simultaneously seeks for maximum diversity and dispersion, introducing the bicriterion approach for anticlustering. Here, we advance the bicriterion approach in several ways. As two minor extensions, we allow that the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices, and that other anticlustering objectives of between-group similarity can be optimized instead of only the diversity. As our main contribution, we present an exact algorithm that returns multiple globally optimal partitions with regard to the dispersion. Despite its theoretical computational hardness, our method scales to rather large data sets (e.g., $N > 5000$ for $K = 3$ groups). By extending existing optimal and heuristic approaches (including BILS), we evaluate how to best obtain high diversity on top of an optimal dispersion. Our analysis reveals an important factor determining the difficulty of the bicriterion anticlustering problem: the level of restriction induced by the maximum dispersion constraint. Depending on the level of restriction, different algorithmic choices are recommended. The original BILS as well as our extensions are available via the free and open source R package `anticlust`. Practical examples are included to illustrate their application.
